Reasonably Programmable Literal Forms

### 2017-07-01 popl18/syntax-popl18.tex
more tweaks to front matter

%% Title information
\title[Reasonably Programmable Syntax]{Reasonably Programmable Syntax}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a few standard data structures, e.g. lists. This practice is \emph{ad hoc} -- there are an unbounded variety of other data structures for which literal forms might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper introduces a system of \emph{typed literal macros (TLMs)} that gives library providers full programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at \emph{any} specified type or parameterized family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 
% Partial parameter application lowers the syntactic cost of this strict style. 
Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program. The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of the spliced terms. The inferred segmentation also informs the hygiene mechanism (previously, only unhygienic macros could freely parse terms out of string literal bodies.) 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules. This is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for an ML- or Scala-like language. % This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that lower the syntactic cost of constructing and pattern matching over values of certain standard types. For example, in Standard ML (SML) and its cousins, the list expression literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. We will return to the concept of context independence again.} List pattern literals are analagous. Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of XHTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Fig. \ref{fig:urweb} gives examples of XHTML literals that ``splice in'' a string expression (Line 1) and another XHTML expression (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This design practice is decidedly \emph{ad hoc} -- it is easy to find other examples of  library constructs that are not privileged with literal forms by these languages, but for which mathematicians, scientists or programmers have invented and adopted specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just XHTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although concerns about literal notation are easy to dismiss as superficial, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL query literals found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of syntactic convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and many other structures. Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are easier than flat string encodings to manipulate programmatically and to reason about compositionally. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-8px}
\label{fig:urweb}
\end{figure}

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with primitive literal forms. Instead, there has been considerable interest in mechanisms that allow library providers to define new literal forms on their own. For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec. \ref{sec:existing-approaches}, can be used to define new literal forms. The problem is that these mechanisms reveal themselves to be quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
Consider, for instance, the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Fig. \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax where a large number of new literal forms, some unfamiliar to the programmer, have been defined using one of these mechanisms.

@@

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently, i.e. to examine the desugaring of the program text, or even the logic that produced that desugaring, defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}. TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) For example, Fig. \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig. \ref{fig:K-dialect}. The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.  TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles absolutely precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}[noitemsep,nolistsep]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax this TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. By examining Fig. \ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way (because they are in green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expressions \li{x} must refer to the binding of \li{x} on Line 2. They cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. (We will see an example of a TLM definition in the next section.) Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

The remainder of this document is organized as follows.
% \begin{itemize}
% \item 
Sec. \ref{sec:setsms} defines a {simple expression TLM} that implements Ur/Web's XHTML literal syntax, and describes the typed expansion process in more detail.
% \item 
Sec. \ref{sec:sptsms} briefly introduces {simple pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above.
% \item 
Sec. \ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two purposes. First, they allow a TLM to operate not just at a single specified type, but over a type- and module-parameterized family of types. We describe a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type. Second, parameters give expansions access to terms other than those explicitly spliced in to the literal body by the client. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application is possible, which lowers the syntactic cost of the strict context independence discipline that TLMs enforce. % Sec. \ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
% \item 
Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of work on macro systems.
% \item 
Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of present limitations and future work.
% \end{itemize}

Due to space limitations, many technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval}, as well as proofs, are relegated to the supplemental material. Examples are written in an emerging dialect of  ML called VerseML that is focused on making (Standard) ML more ergonomic. However, the essence of the mechanism is not ML-specific -- it should be straightforward to adapt it to other languages that take a similar approach to types and binding (e.g. Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}. Languages that have a disciplined binding structure but that lack a rich static type structure, e.g. Racket and other LISP derivatives, would also benefit from TLMs -- they can apply the results of this paper by deploying the usual type theorist's trick of viewing such a language as ``{unityped}'' \cite{pfpl,scott1980lambda}. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

We will begin in this section with another example that informally introduces \emph{simple expression TLMs} (seTLMs), i.e. TLMs for generating expressions of a single specified type.

Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TLM named 
\li{$html} 
 to \emph{generalized literal forms} delimited by backticks. TLM names are prefixed by \li{$} 
 to clearly them from variable identifiers. The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase. 
\begin{figure}[ht!]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
\end{lstlisting}
\vspace{-10px}
\caption{Two examples of a TLM being applied to a generalized literal form. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase. 
We assume strings are primitive.}
\vspace{-10px}
\label{fig:first-tsm-example}
\end{figure}

>>>>

%% Title information
\title[Reasonably Programmable Literal Forms]{Reasonably Programmable Literal Forms}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
General-purpose programming languages typically provide literal notation for only a small number of standard data structures, like lists. This language design practice is decidedly \emph{ad hoc}: there are many more data structures for which literal notation might also be useful (consider, for example,  maps, regexes, HTML data, SQL queries, syntax trees and chemical structures.) Availability of literal notation is far from a frivolous concern: studies suggest that literal notation can have a non-trivial impact on software quality by removing the temptation to use string encodings, which are vectors for injection attacks and cause other difficulties. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.
This paper details a mechanism for decentralizing control over literal notation: \emph{typed literal macros (TLMs)}. TLMs give library providers programmatic control over the parsing and expansion of expressions and patterns of \emph{generalized literal form} at {a} specified type or parameteric family of types. % The mechanism is {strictly hygienic}, meaning that 1) the expansion must be context-independent; and 2)  splicing is capture avoiding. 
% Partial parameter application lowers the syntactic cost of this strict style. 
Compared to existing approaches, including syntax extension systems and macro-based quotation systems, TLMs are uniquely \emph{reasonable}: TLM providers can reason modularly about syntactic determinism, and TLM clients can reason abstractly about the type and binding structure of the program. The system only needs to convey to clients (via secondary notation) the inferred \emph{segmentation} of each literal, which gives the locations of spliced terms. Maintaining information about segmentation supports the novel TLM hygiene mechanism, i.e. the mechanism that ensures that clients can reason abstractly about binding (previously, only unhygienic macros could freely splice terms out of string literal bodies.) 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
This paper establishes these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules. This calculus is the  first detailed type-theoretic account of a typed, hygienic term-rewriting macro system of any design for a language with these features of ML or Scala. % This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to build in a few \emph{literal forms} that decrease the syntactic cost of constructing and pattern matching over values of certain standard types. For example, in Standard ML (SML) and its cousins, a list expression literal like \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} identify the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. We will return to the concept of context independence again.} Another example is found in Ur/Web's surface syntax, which extends Ur's surface syntax with expression and pattern literal forms for encodings of HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Fig. \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2). 
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This language design practice is decidedly \emph{ad hoc} -- it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have elsewhere invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. Consider, for example, that 1) clients of a ``collections'' library might like not just list literals, but also matrix, set and map literals following standard mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web does not provide); 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like composite chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although concerns about ``missing'' literal notation are easy to dismiss as superficial or frivolous, the reality is that literal forms, or the absence thereof, have had a substantial influence on the quality of software that is written today. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal forms for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{LINQ}\todo{find citation}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of syntactic convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and many other data structures. Literal forms at these types, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-10px}
\caption{HTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-8px}
\label{fig:urweb}
\end{figure}

Of course, it would not be sensible to ask general-purpose language designers to accomodate all known notations with built-in literal forms. Instead, there has been considerable interest in mechanisms that allow library providers to define new literal notation on their own. For example, general syntax extension mechanisms, including Camlp4 \cite{ocaml-manual}, SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and others that we will return to in Sec. \ref{sec:existing-approaches}, can be used to define new literal forms. The problem that specifically motivates this paper is that these mechanisms are quite \emph{unreasonable} when ``programming in the large'' \cite{DeRemer76}.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''
Consider, for instance, the perspective of a programmer attempting to comprehend (i.e. reason about) the program text in Fig. \ref{fig:K-dialect}, which is written in a dialect of ML's (or Scala's) syntax, constructed by one of these mechanisms, where a large number of new literal forms have been defined.

@@

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently, i.e. to examine the desugaring of the program text, or even the logic that produced that desugaring, to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML and Scala -- the binding discipline of the language ensures that programmers can reason lexically about binding, and types mediate abstraction over function and module implementations. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

This paper formally details a more reasonable mechanism for defining new literal forms: \emph{typed literal macros (TLMs)}. TLMs give library providers full programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) For example, Fig. \ref{fig:K-tsm-example} applies a TLM, \li{$kquery}, to a backtick-delimited generalized literal form to express the ``unreasonable'' K query example from Fig. \ref{fig:K-dialect} just discussed. The applied TLM parses the literal body and generates an expansion statically, during a semantic phase called typed expansion, which generalizes the usual typing phase.  TLMs come equipped with uniquely powerful abstract reasoning principles. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop some intuitions: the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}[noitemsep,nolistsep]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, is solely responsible for parsing and expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider can modularly establish that the syntax this TLM implements is deterministic.
\item \textbf{(Segmentation)} The output that a TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. By examining Fig. \ref{fig:K-tsm-example}, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K query language (because they are in green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

The remainder of this document is organized as follows.
% \begin{itemize}
% \item 
Sec. \ref{sec:setsms} defines another {simple expression TLM} -- one that implements Ur/Web-style HTML literal syntax (Fig. \ref{fig:urweb}) -- and uses it as a running example to informally detail how TLMs work from start to finish.
% \item 
Sec. \ref{sec:sptsms} briefly introduces {pattern TLMs}, i.e. TLMs that generate patterns. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec. \ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
% \item 
Sec. \ref{sec:ptsms} then adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that take type and module parameters. Parameters serve two key purposes. First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types. We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type. Second, parameters give expansions access to terms other than those explicitly spliced in to the literal body by the client. %This is particularly convenient for working with abstract types \emph{a la} ML. 
Partial parameter application is possible. Together, this lowers the syntactic cost of the strict context independence discipline that TLMs enforce. % Sec. \ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in  more detail, and also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.
% \item 
Sec. \ref{sec:existing-approaches} compares TLMs to related work, including recent related work by \citet{TSLs,sac15}, and the long history of related work on macro systems.
% \item 
Sec. \ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a brief discussion of the limitations of TLMs, along with some suggestions for future work.
% \end{itemize}

Due to space limitations, certain technical details of the calculi defined in this paper, as well as proofs, are relegated to the supplemental material. The examples in this paper are written in an emerging dialect of  ML called VerseML that is focused on improving the ergonomics of ML. However, the essential ideas are not ML-specific -- it should be possible to adapt TLMs to other languages that take a similar approach to types and binding (e.g. Haskell, Scala and others.) When a design choice is ML-specific, we will make note of it\todo{do this}. Languages that have a disciplined binding structure but that lack a rich static type structure, e.g. Racket and other LISP derivatives, might also benefit from TLMs -- they can apply the results of this paper by deploying the usual trick of reducing the language to one that is ``{unityped}'' \cite{pfpl,scott1980lambda}.%  We say more about LISP macros in Sec. \ref{sec:existing-approaches}.% For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

@@

This section informally details the TLM mechanism. It will suffice in this section to consider only \emph{simple expression TLMs (seTLMs)}, i.e. TLMs for abbreviating expressions of a single specified type. In particular, this section is organized around an seTLM named \li{$html} that implements Ur/Web-style HTML literals, which abbreviate expressions of the type \li{html} defined as follows:
\begin{lstlisting}[numbers=none]
  type html = BodyElement of list(html_attr) * list(html)
            | H1Element of list(html_attr) * list(html) 
            | (* ... *) | TextNode of string 
\end{lstlisting}
The type \li{html_attr} is straightforward and the type function \li{list} is standard, so neither are shown. In VerseML, \li{string}s are primitive (though this is not essential, see footnote \todo{reference}1 below.)

Using this TLM, we can express the Ur/Web program from Fig. \ref{fig:urweb} as shown in Fig. \ref{fig:first-tsm-example}: on each line, we apply 
\li{$html} 
 to a \emph{generalized literal form} delimited by backticks. TLM names are prefixed by \li{$} 
 to clearly them from variable identifiers. The semantics delegates control over the parsing and expansion of each literal body to the applied TLM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase. 
\begin{figure}[ht!]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
\end{lstlisting}
\vspace{-10px}
\caption{Two examples of a TLM being applied to a generalized literal form. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase. 
We assume strings are primitive.}
\vspace{-10px}
\label{fig:first-tsm-example}
\end{figure}
